### 2010-09-14 Thesis/Chapter10.tex
- Update 14 September -
Chapter 7, 9 and 10 finished and sent for review



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@175 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Physics-based reconstruction using shell elements}
\label{chap10}
\begin{shortAbstract}
A short abstract for the upcoming chapter \\
\OC{MICCAI 2010}
\end{shortAbstract}

@@

Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved. In the following, we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the object we want to simulate (via a Marching Cube algorithm for instance, \cite{Lorensen87}). Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry. For this reason, we need a technique to simplify the high resolution mesh of the object to be simulated.

\TODO{state of the art to be added... }

@@

However, we need to keep in mind that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure. In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell. We propose to approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.

The first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in \cite{Meshlab}. The algorithm tries as much as possible to preserve mesh boundaries and generates high quality triangular elements. We then apply a heuristic method derived from the work of \cite{Saupin07} with tetrahedral meshes based on simple geometrical rules. For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycentre of the three centres of mass of those triangles. This technique locally smoothes the surface of the mesh while converging towards the desired high resolution mesh.

@@

The efficiency of the method is illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures. We also present preliminary results on the simulation of an angioplasty procedure.

@@

\TODO{Did we answer our problematic? Limitations? }

>>>>

\chapter{Physics-based reconstruction using shell elements}
\label{chap10}
\begin{shortAbstract}
In the previous chapter, we proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach. This simple shell element can efficiently handle both membrane and offplane forces (bending and twist). The validity of our approach has been demonstrated though comparison with theoretical results. It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation. These preliminary results are very encouraging and show the potential of such models. Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels. However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches. In the next section, we propose to study how to approximate the surface of anatomical structures with (curved) shell elements. This work was the object of one publication \citep{Comas2010b}.
\end{shortAbstract}

@@

Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved. Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry. Literature about volumic mesh generation algorithms is fairly dense. However, there are only a few that are concerned about the generation of meshes over curved surfaces. One of the most widely used techniques for the creation of surface meshes is the plane to surface transformation method \citep{Zienkiewicz71}, mesh is first generated on a two-dimensional domain and then mapped onto the surface. If this method gives reasonably good meshes on smooth surfaces, the results are usually quite poor with more complex curved surfaces. The finite elements may be generated directly on the curved surfaces based on the advancing front technique \citep{Lo85,Lau96}. The main issue with this approach is that an analytical description of the geometry is needed, which is not the case in medical simulation. Another method consists of using an a priori error estimator to build an adaptive mesh generation \citep{Baumann97}. However, the tolerance of this indicator should be chosen depending on the desired accuracy of the finite element solution, and therefore requires some knowledge about the problem in order to choose an effective tolerance. \cite{Bechet02} start from an existing triangular mesh created with a CAD (computer-aided design) software and refine and smooth the mesh based on element quality and surface curvature. While all those methods allow their authors to get satisfactory meshes over curved surfaces according to their needs, they all make use of flat elements to mesh geometries and do not generate actual shells.

In fact, it is interesting to note that this process of meshing a curved surface is similar to the reconstruction of the surface of objects in computer vision. Indeed, calculating curvature maps of 3D surfaces represented as digitised data (point clouds or triangulated meshes) has been extensively studied. One of the most common approach is to use continuous surface patches \citep{Kolb95,Douros02}. The surface is locally approximated by an analytic representation called surface patch, usually chosen to be an implicit surface. \cite{Tang99} presented an elegant and non noise-sensitive approach that infers sign and direction of principal curvatures directly from the input and the authors used this information for coherent surface and curve extraction from 3D data.

However, we need to keep in mind that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure. In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell. By keeping this in mind, we propose an approach to mesh a curved surface based on the same polynomial shape function \eqref{chap9:eq-deflection} used in our co-rotational shell FEM.

@@

In the following, we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the object we want to simulate (via a Marching Cube algorithm for instance, \cite{Lorensen87}). We propose a technique that can simplify the high resolution mesh of the object to be simulated and approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation. Thus, the first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in \cite{Meshlab}. MeshLab is an open source program for the processing and editing of unstructured 3D triangular meshes. The algorithm tries as much as possible to preserve mesh boundaries and generates high quality triangular elements. We then apply a heuristic method derived from the work of \cite{Saupin07} with tetrahedral meshes based on simple geometrical rules. For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycentre of the three centres of mass of those triangles. This technique locally smoothes the surface of the mesh while converging towards the desired high resolution mesh.

@@

We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy. The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects (like an intra-ocular implant). We also presented preliminary results on the coupling between solid (tetrahedra) and thin objects (shells) for the advanced modelling of anatomical structures via the simulation of an angioplasty procedure.

@@

The finite element procedure that we propose is largely developed based on physical understanding without the use of mathematical shell theories. Indeed, our shell FEM was obtained by simply superimposing a plane stress and bending energies. According to \cite{Chapelle03}, this type of approach has a limited level of accuracy. The effective study of shell structures clearly requires a deep physical understanding of shell structural behaviours. The complexity of the physical behaviours of shells requires advanced mathematical analyses from theories that are still in development. The interested reader may refer to the book of \cite{Chapelle03} for more details on shell theory. If the range of application of our shell formulation might be somewhat limited, we did not find out any inconsistencies through our tests. We also need to keep in mind that the accuracy required in medical simulation is not the same than the one demanded in structural analyses for instance. Moreover, the additional complexity of a shell finite element formulation based on true shell theory may also very well jeopardise our real-time constraints. Therefore, we do not think that the use of this simple shell formulation significantly diminishes our contribution to the field of medical simulation. Indeed, to our knowledge, our co-rotational shell FEM is the first description of a shell finite element model applied to simulating the deformation of thin anatomical structures.

Nevertheless, a few improvements are conceivable. Large strain could be allowed by using a non-linear strain tensor and more complex material laws (non-linear, viscoelasticity, etc.) could also be added to our shell formulation. However, those changes are substantial and their implementations are not straightforward. In addition, it is not clear how the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true shell theory (where membrane and bending energies are not separated). The answer may depend on the type of constraints applied on the object. Also (as already mentioned section \ref{chap9:GPU}), the real-time capacity of our algorithm could be further improved by implementing our shell FEM on GPU. Although an optimal parallel re-implementation is always challenging, we have already reviewed why there was no major obstacle to a GPU implementation.

Although we were able to mesh complex anatomical structures with a fairly small number of shell elements, our approach suffers from two major drawbacks. In our current implementation, if the error measured with the Hausdorff distance is greater than a threshold (which is also difficult to choose), all triangular elements are subdivised into 4 new triangles and each of the new vertices is moved towards the targeted geometry based on our heuristic algorithm. Obviously, this is not optimal as the error between the two meshes may very well be local and therefore the use of a global mesh refinement may be very limited. The second drawbrack is that we also do not check the quality of the elements during our mesh process. If some elements are stretched from the heavy decimation, these elements are not corrected during the process of remeshing. Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements. One solution for the latter issue would be to enhance our algorithm with a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object). This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall. Yet, meshing a surface with curved elements is a very complex problem and still an active area of research. In contrast to all previous geometrical techniques, we suggested a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.

### 2010-09-14 Thesis/Chapter7.tex
- Update 14 September -
Chapter 7, 9 and 10 finished and sent for review



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@175 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Modelling the deformation of hollow objects in real-time}
\label{chap7}
\begin{shortAbstract}
A short abstract for the upcoming chapter
\end{shortAbstract}

\section{Introduction: the problematic}
The human body is composed of various deformable anatomical structures. A key challenge of soft-tissue modelling is the variousness of the mechanical behaviours. The shape and the internal structure may be very different from one organ to another. If you take the liver for instance, it is a solid organ composed of four lobes, each one made up of units called lobules. Each lobule consists of a central vein surrounded by liver cells and constitutes the end of a dense network of blood vessels. This particular internal structure has a strong influence on the mechanical properties of the liver. In contrast, the colon is a long and hollow organ. It consists of four sections which all have different shapes and material properties. Consequently, it seems unrealistic to use a unique model for all tissues. Yet, most of previous works focus on volumetric models that are able to capture the behaviour of solid organs. In fact in the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures whose volume is negligible compared to their surface area. Examples include hollow structures, such as the wall of blood vessels, or membranes, such as the Glisson?s capsule surrounding the liver. It is also of particular interest to us for modelling the colon in our colonoscopy simulator and for simulating the deployment of the implant in cataract surgery.

The first idea that may come to mind for modelling thin structures is to apply the same methods than with solid objects. For instance, \cite{Holzapfel02} proposed a very complex 3-layer model of an aterial wall for simulating an angioplasty procedure. Their model is non-linear and anisotropic along fibers placed according to histology. They also model diseased part of the vessel through a hardening parameter which adds plasticity. However, they discretised their geometric model with eight-node brick elements. Another example is given by \cite{Aloisio04} who assume that the blood flow in the vessels may be considered as incompressible and the artery consequently becomes a solid body and can be modelled as a solid deformable object. \TODO{why is this not appropriate? High-order elements could do the trick to follow the curvature...} Hence, specific models for thin objects were developed.

>>>>

\chapter{Modelling the deformation of hollow objects in real-time}
\label{chap7}
\begin{shortAbstract}
The human body is composed of various deformable anatomical structures and finding appropriate soft-tissue models for all these type of structures is challenging. In the previous two chapters, we described an efficient non-linear FEM formulation for simulating the deformation of solid organs (like brain, liver, prostate etc.). But there is also a need to simulate the deformation of hollow organs, whose volume is negligible compared to their surface area, such as colon, blood vessels, stomach etc. The solid models previously mentioned are not appropriate and specific approaches are required for these thin anatomical structures. In this chapter, we will review diverse techniques applied for the simulation of thin objects. These different approaches may be grouped into three main categories: (1) mass-spring models, (2) approaches which derive a bending energy from geometrical considerations and (3) methods based on the equations of continuum mechanics. Although very few models have been proposed for simulating the deformation of thin anatomical structures in the field of medical simulation, we will try to emphasize medical applications whenever possible. 
\end{shortAbstract}

\section{Introduction: the problematic}
The human body is composed of various deformable anatomical structures. A key challenge of soft-tissue modelling is the variousness of the mechanical behaviours. The shape and the internal structure may be very different from one organ to another. If you take the liver for instance, it is a solid organ composed of four lobes, each one made up of units called lobules. Each lobule consists of a central vein surrounded by liver cells and constitutes the end of a dense network of blood vessels. This particular internal structure has a strong influence on the mechanical properties of the liver. In contrast, the colon is a long and hollow organ. It consists of four sections which all have different shapes and material properties. Consequently, it seems unrealistic to use a unique model for all tissues. Yet, most of previous works focus on volumetric models that are able to capture the behaviour of solid organs. In fact in the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures whose volume is negligible compared to their surface area. Examples include hollow structures, such as the wall of blood vessels, or membranes, such as the Glisson's capsule surrounding the liver. It is also of particular interest to us for modelling the colon in our colonoscopy simulator and for simulating the deployment of the implant in cataract surgery.

The first idea that may come to mind for modelling thin structures is to apply the same methods than with solid objects. For instance, \cite{Holzapfel02} proposed a very complex 3-layer model of an aterial wall for simulating an angioplasty procedure. Their model is non-linear and anisotropic along fibers placed according to histology. They also model diseased part of the vessel through a hardening parameter which adds plasticity. However, they discretised their geometric model with eight-node brick elements. Another example is given by \cite{Aloisio04} who assume that the blood flow in the vessels may be considered as incompressible and the artery consequently becomes a solid body and can be modelled as a solid deformable object. \TODO{why is this not appropriate? High-order elements could do the trick of following the curvature...} Hence, specific models for thin objects were developed.

### 2010-09-14 Thesis/Chapter9.tex
- Update 14 September -
Chapter 7, 9 and 10 finished and sent for review



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@175 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{A co-rotational triangular shell FEM model}
\label{chap9}
\begin{shortAbstract}
A short abstract for the upcoming chapter \\
\OC{ISBMS 2010}
\end{shortAbstract}

@@

\subsection{Possible extension: parallel implementation on GPU}
If we look closely at all the steps required to compute the local stiffness matrix for each element (see section \ref{chap9:summary} for a summary), we can notice that the operations carried out for a given shell element are independent of the other elements. This feature makes our co-rotational shell FEM implementation highly parallelisable. One potential problem however, is the assembly of the global stiffness matrix. Indeed, in a parallel implementation, many elements are processed concurrently. Inevitably, two elements sharing a node may be processed in the same time and may therefore try to both write their element contribution into the exact same location in the global stiffness matrix, and therefore the same location in GPU's memory. As with the TLED implementation described in chapter \ref{chap6}, there is two solutions. The first one would be to re-order the elements organisation so that no concurrent writes will occur. Unfortunately, this re-organisation of the element reduces the number of elements that can be processed in parallel. The second solution would be to take advantage of a new GPU NVIDIA architecture called Fermi, which added the capability atomic writes for floats. We recall that the write operation is said to be atomic in the sense that it is guaranteed to be performed without interference from other threads. In other words, the concurrent writes are simply serialised in hardware. Obviously, an additional overhead is to be expected.

@@

We compared our model with some theoretical results reported by \cite{Zhongnian86} to assess its quality in modelling bending. The test that we carried out uses a square shape mesh clamped on all four edges. A uniform load is then applied on the square and the maximum deflection $z_{max}$ at the centre can be calculated. Several simulations are performed with increasing load values $q$ (ranging from $1$ to $5 N/m^2$) and the following parameters were used: 
\begin{itemize}
 \item Young Modulus $E = 1.092 \times 10^6\,$Pa
 \item Poisson's ratio $\nu = 0.3$
 \item edge of the square $L = 10\,m$
 \item thickness $h = 0.1\,m$
 \item pressure q is a uniformly distributed load per unit area
\end{itemize}

@@

\section{Discussion}

We proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach. This simple shell element can efficiently handle both membrane, bending and twist forces. The validity of our approach has been demonstrated though comparison with theoretical results. It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation. These preliminary results are very encouraging and show the potential of such models. Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels. However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches. In the next section, we propose to study how to approximate the surface of anatomical structures with shell elements.

>>>>

\chapter{A co-rotational triangular shell FEM model}
\label{chap9}
\begin{shortAbstract}
In the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures. In fact, to our knowledge, no model based on the equations of continuum mechanics was described and applied to simulating the deformation of objects. In this chapter, we propose to rely on continuum mechanics to accurately describe the complex deformations of the implant. Our approach extends the co-rotational method used in finite element analysis of in-plane deformations to incorporate a bending energy. We will first describe the model in details before discussing its implementation in the open source framework SOFA. Lastly, this co-rotational shell FEM was applied to the simulation of the lens deployment during catarct surgery. This simulation also accounts for the complex contacts that take place during the injection phase. This work was the object of two publications \citep{Comas2010a,Comas2010b}.
\end{shortAbstract}

@@

\subsection{Possible extension: parallel implementation on GPU}	\label{chap9:GPU}
If we look closely at all the steps required to compute the local stiffness matrix for each element (see section \ref{chap9:summary} for a summary), we can notice that the operations carried out for a given shell element are independent of the other elements. This feature makes our co-rotational shell FEM implementation highly parallelisable. One potential problem however, is the assembly of the global stiffness matrix. Indeed, in a parallel implementation, many elements are processed concurrently. Inevitably, two elements sharing a node may be processed in the same time and may therefore try to both write their element contribution into the exact same location in the global stiffness matrix, and therefore the same location in GPU's memory. As with the TLED implementation described in chapter \ref{chap6}, there is two solutions. The first one would be to re-order the elements organisation so that no concurrent writes will occur. Unfortunately, this re-organisation of the element reduces the number of elements that can be processed in parallel. The second solution would be to take advantage of a new GPU NVIDIA architecture called Fermi, which added the capability atomic writes for floats. We recall that the write operation is said to be atomic in the sense that it is guaranteed to be performed without interference from other threads. In other words, the concurrent writes are simply serialised in hardware. Obviously, an additional overhead is to be expected.

@@

We compared our model with some theoretical results reported by \cite{Zhongnian86} to assess its quality in modelling bending. The test that we carried out uses a square shape mesh clamped on all four edges. A uniform load is then applied on the square and the maximum deflection $z_{max}$ at the centre can be calculated. Several simulations are performed with increasing load values $q$ (ranging from $1$ to 5 N/m$^2$) and the following parameters were used: 
\begin{itemize}
 \item Young Modulus $E = 1.092 \times 10^6\,$Pa
 \item Poisson's ratio $\nu = 0.3$
 \item edge of the square $L = 10\,$m
 \item thickness $h = 0.1\,$m
 \item pressure q is a uniformly distributed load per unit area
\end{itemize}

@@
